Formula translation in C++, Python and modern Fortran: a~case~study~of~the~language~choice tradeoffs

### 2013-07-26 paper/doc/paper.tex
fixing the donor-cell parenthesis issue; paper.tex updated befor Sci. Prog. submission

    \author[1]{Sylwester Arabas\footnote{Corresponence to: S. Arabas -- sarabas@igf.fuw.edu.pl -- 0048225546827 -- University of Warsaw, Pasteura 7, 02-093 Warszawa, Poland}}
    \author[1]{Dorota Jarecka}
    \author[1]{Anna Jaruga}
    \author[2]{Maciej Fijałkowski}

    \affil[1]{Institute of Geophysics, Faculty of Physics, University of Warsaw}
    \affil[2]{PyPy Team}

@@

  Object oriented programming (OOP) {\em ''has become recognised as the almost unique successful 
    paradigm for creating complex software''} \citep[][Sec.~1.3]{Press_et_al_2007}.
  It is intriguing that, while the quoted statement comes from the very book subtitled 
   {\em The Art of Scientific Computing}, hardly any (if not none) of the currently operational 
    weather and climate prediction systems - flagship examples of complex scientific software - 
    make extensive use of OOP techniques\footnote{
  Fortran has been the language of choice in oceanic \citep{Griffies_et_al_2000}, 
    weather-prediction \citep{Sundberg_2009} and Earth system \citep{Legutke_2012} modelling, 
    and none of its 20-century editions were object-oriented languages \citep[see e.g.][for discussion]{Norton_et_al_2007}}.

  Application of OOP techniques in development of numerical modelling software may help to:
  \begin{enumerate}[label=(\roman*), leftmargin=*, widest=ii]
    \item{maintain modularity and separation of program logic layers (e.g. separation of
      numerical algorithms, parallelisation mechanisms, data input/output, error handling and
      the description of physical processes); and}
    \item{{\bf shorten and simplify the source code and improve its readability by reproducing within 
      the program logic the mathematical notation used in the literature}.}
  \end{enumerate}
  The first application is attainable, yet arguably cumbersome, with procedural programming.
  The latter, virtually impossible to obtain with procedural programming, is the focus of this paper.
  The importance of reproducing the mathematical notation in the code lays primarily
    in the fact that code readability and brevity \citep[and not the language choice itself,][]{Seeley_2004}
    significantly contribute to code maintainability \citep{Wilson_et_al_2012}.

@@

  All three programs include an equally structured implementation of the two-dimensional version of the 
    Multidimensional Positive Definite Advective Transport Algorithm \citep[MPDATA;][]{Smolarkiewicz_1984}.
  MPDATA is an example of a numerical procedure used in weather, climate, ocean and solar simulation systems
    \citep[e.g.][respectively]{Ziemianski_et_al_2011,Abiodun_et_al_2011,Ezer_et_al_2002,Charbonneau_and_Smolarkiewicz_2013}.
  The basic MPDATA scheme presented herein is complex enough to contain
    a wide range of mathematical abstractions that can be represented using OOP constructs, 
    yet it is simple enough to allow inclusion of the entire source code within the paper text.
  All relevant MPDATA formul\ae~are given in the text alongside corresponding code fragments
    allowing comparison of the relevant syntax with the mathematical notation.
  In compliance with the article focus, these formul\ae~are presented without derivation or detailed discussion
    (see~\citep{Smolarkiewicz_2006} for a recent review of MPDATA-based techniques including an introductory description 
    of the algorithm and an exhaustive list of references).

  The paper is structured as follows.
  In section~\ref{sec:impl} we introduce the ''formula translation'' part of the three implementations
    briefly describing the algorithm itself and
    discussing where and how the OOP techniques were applied in its implementation.
  The remaining part of the implementations -- the solver logic -- is
    presented in appendix~\ref{sec:app}.
  Section~\ref{sec:perf} covers performance evaluation of the three implementations.
  Section~\ref{sec:tradeoffs} covers discussion of the tradeoffs of the programming language choice.
  Section~\ref{sec:concl} closes the article with a brief summary.

@@

  \section{Implementation: the formul\ae}\label{sec:impl}

@@

  In NumPy in order to prevent copying of array data during slicing one needs to operate on the
    so-called array views.
  Array views are obtained when indexing the arrays with objects of the Python's
    built-it \prog{slice} type (or tuples of such objects in case of multi-dimensional arrays).
  Python forbids overloading of operators of built-in types such as \prog{slices}, 
    and does not define addition/subtraction operators for \prog{slice} and \prog{int} pairs.
  Consequently, a custom logic has to be defined not only for fractional indexing,
    but also for shifting the slices by integer intervals ($i\pm1$).
  It is implemented here by declaring a \prog{shift} class with the adequate operator overloads:
  \codepyt{code-pyt-listings.py}{listing03}{listing04}{1}
    and two instances of it to represent unity and half
    in expressions like \prog{i+one}, \prog{i+hlf}, where \prog{i} is an instance of \prog{slice}%
    \footnote{\label{fnt:slice}One could argue that not using an own implementation of a slice-representing class
    in NumPy is a design flaw -- being able to modify behaviour of a hypothetical numpy.slice class 
    through inheritance would allow to implement the same behaviour as obtained in listing P.3 without the need to represent 
    the unity as a separate object}:
  \codepyt{code-pyt-listings.py}{listing04}{listing05}{1}

@@

  In the Fortran implementation \prog{pi()} returns a pointer to the array elements specified
    by \prog{i} and \prog{j} interpreted as (i,j) or
    (j,i) depending on the value of the argument \prog{d}.
  In addition to \prog{pi()}, a helper \prog{span()} function returning the 
    length of one of the vectors passed as argument is defined:
  \codefor{code-for-listings.f}{listing04}{listing05}{1}
  The \prog{span()} function is used to shorten the declarations of arrays to be returned 
    from functions in the Fortran implementation (see listings F.7 and F.12--F.15).

@@

@@

  If the denominator in equations \ref{eq:A} or \ref{eq:B} equals zero for a given {\em i} and {\em j}, 
    the corresponding $A_{[i,j]}$ and $B_{[i,j]}$ are set to zero what may be conveniently 
    represented with the \prog{where} construct in all three considered languages:
  \codecpp{code-cpp-listings.hpp}{listing10}{listing11}{1}
  \codepyt{code-pyt-listings.py}{listing09}{listing10}{1}
  \codefor{code-for-listings.f}{listing10}{listing11}{1}
  \codefor{code-for-listings.f}{listing11}{listing12}{1}

@@

@@

  In addition to the standard Python implementation CPython,
    the Python code was tested with PyPy.
  PyPy is an alternative implementation of Python featuring a just-in-time compiler. 
  PyPy includes an experimental partial reimplementation of NumPy that compiles NumPy expressions into native assembler.
  Thanks to employment of lazy evaluation of array expressions (cf. Sect. \ref{sec:array})
    PyPy allows to eliminate the use of temporary matrices for storing intermediate results,
    and to perform multiple operations on the arrays within a single array index traversal
    \footnote{Lazy evaluation available in PyPy 1.9 has been temporarily removed from PyPy during a refactoring of
      the code. It'll be reinstantiated in the codebase as soon as possible, but past PyPy 2.0 release}.
  Consequently, PyPy allows to overcome the same performance-limiting factors as those addressed by Blitz++, although 
    the underlying mechanisms are different.
  In contrast to other solutions for improving performance of NumPy-based codes such as
    Cython\footnote{see \url{http://cython.org}}, 
    numexpr\footnote{see \url{http://code.google.com/p/numexpr/}} or 
    Numba\footnote{see \url{http://numba.pydata.org/}}, 
    PyPy does not require any modifications to the code.
  Thus, PyPy may serve as a drop-in replacement for CPython ready to be used with 
    previously-developed codes.

  The same set of tests was run with all four set-ups.
  Each test set consisted of 16 program runs.
  The test programs are analogous to the example code presented in section~\ref{sec:example}.
  The tests were run with different grid sizes ranging from 64$\times$64 to 2048$\times$2048.
  The Gaussian impulse was advected for $nt=2^{24}/(nx\cdot ny)$ timesteps ($2^{24}$ chosen arbitrarily), 
    in order to assure comparable timing accuracy for all grid sizes.
  Three MPDATA iterations were used (i.e. two corrective steps).
  The initial condition was loaded from a text file, and the final values were compared at the end of the test
    with values loaded from another text file assuring the same results were obtained with all four set-ups.
  The tests were run multiple times; program start-up, data loading, and output verification times were
    subtracted from the reported values (see caption of Figure~\ref{fig:cpu-eyrie} for details).

@@

  The CPU time statistics presented in Figures \ref{fig:cpu-eyrie} and \ref{fig:cpu-skua} reveal
    minor differences between results obtained with the two different processors.
  Presented results lead to the following observations
    (where by referring to language names, only the results obtained with the herein considered
     program codes, and software/hardware configurations are meant):
  \begin{itemize}
    \item{Fortran gives shortest execution times for any domain size;}
    \item{C++ execution times are less than twice those of Fortran for grids larger than 
      256$\times$256;}
    \item{CPython requires from around 4 to almost 10 times more CPU time than Fortran depending on the grid size;}
    \item{PyPy execution times are in most cases closer to C++ than to CPython.}
  \end{itemize}
  The support for OOP features in gfortran, the NumPy support in PyPy, and the relevant optimisation
    mechanisms in GCC are still in active development and hence the performance with some of the set-ups may 
    likely change with newer versions of these packages.

@@

  Three issues specific to Fortran that 
    resulted in employment of a more repetitive or cumbersome syntax than in C++ or Python
    were observed:
  \begin{itemize}
    \item{Fortran does not feature a mechanism allowing to reuse a single piece of code (algorithm)
      with different data types (compare e.g. listings C.15, P.14 and F.17) such as
      templates in C++ and the so-called duck typing in Python;}
    \item{Fortran does not allow function calls to appear on the left hand side
      of assignment (see e.g. how the \prog{ptr} pointers were used as a workaround in the \prog{cyclic\_fill\_halos}
      method in listing F.20);}
    \item{Fortran lacks support for arrays of arrays (cf. Sect. \ref{sec:sequence}).}
  \end{itemize}
  Interestingly, the limitation in extendability via inheritance was found to
    exist partially in NumPy as well (see footnote \ref{fnt:slice}).
  The lack of a counterpart in Fortran to the C++ template mechanism was identified in
    \citep{Cary_et_al_1997}
    as one of the key deficiencies of Fortran when compared with C++ in context 
    of applicability to object-oriented scientific programming.

@@

  With both Fortran and Python, the memory footprint caused by employment
    of temporary objects in array arithmetics is dependant on compiler choice or
    the level of optimisations.
  In contrast, Blitz++ ensures temporary-array-free computations by design
    \citep{Veldhuizen_et_al_1997} avoiding unintentional performance loss.

  \subsection{User- and code-base}

@@

  Finally, the three languages differ as well with regard to availability of 
    mechanisms for leveraging shared-memory parallelisation (e.g. with multi-core processors).
  GCC supports OpenMP with Fortran and C++.
  The CPython and PyPy implementations of Python do not offer any
    built-in solution for multi-threading.

@@

  The key conclusion is that all considered language/library/compiler
    set-ups offer possibilities for using OOP to compactly 
    represent the mathematical abstractions within the program code. 
  This creates the potential to improve code readability and brevity,
  \begin{itemize}
    \item{contributing to its 
      auditability, indispensable for credible and reproducible research in computational science 
      \citep{Post_et_al_2005, Merali_et_al_2010, Stodden_et_al_2012}; and
    }
    \item{helping to keep the programs maintainable and avoiding accumulation of the code 
      debt\footnote{See \citet{Buschmann_2011} for discussion of technical/code debt.} 
      that besets scientific software in such domains as climate modelling
      \citep{Freeman_et_al_2010}.
    }
  \end{itemize}
  \noindent 
  The performance evaluation revealed that:
    \begin{itemize}
      \item{the Fortran set-up offered shortest execution times,}
      \item{it took the C++ set-up less than twice longer to compute than Fortran,}
      \item{C++ and PyPy set-ups offered significantly smaller memory consumption 
        than Fortran and CPython for larger domains,}
      \item{the PyPy set-up was roughly twice slower than C++ and up to twice faster than CPython.}
    \end{itemize}
  The three equally-structured implementations required ca. 200, 300, and 500 lines of code 
    in Python, C++ and Fortran, respectively.

@@

  \section{Implementation: the solvers}\label{sec:app}
  \subsection{Halo regions}

@@

  The \prog{advop()} method controlls the MPDATA iterations within one timestep.
  The first (step = 0) iteration of MPDATA is an unmodified donor-cell step.
  Subsequent iterations begin with calculation of the antidiffusive Courant fields using 
    formula \ref{eq:antidiff1}.
  In order to calculate values spanning an (i\mhlf~... i+\phlf) range using a formula
    for $C_{[i+\phlf,\ldots]}$ only, the formula is evaluated using extended
    index ranges \prog{im} and \prog{jm}.
  In the second (step=1) iteration the uncorrected Courant field (\prog{C\_unco}) points
    to the original \prog{C} field, and the antidiffusive Courant field is written into
    \prog{C\_corr} which points to \prog{tmp[1]}.
  In the third (step=2) iteration \prog{C\_unco} points to \prog{tmp[1]} while \prog{C\_corr}
    points to \prog{tmp[0]}.
  In subsequent iterations \prog{tmp[0]} and \prog{tmp[1]} are alternately swapped.

@@

  \subsection{Usage example}\label{sec:example}

@@

  \small

@@

  \begin{figure}[bh!]
    \caption{\label{fig:mpdata}
      Plot generated by the program given in listing~C.20.
      The top panel shows initial signal shape (at time t=0).
      The subsequent panels show snapshots of the advected field after 75 timesteps
        from three different simulations: donorcell (or 1 MPDATA iteration), MPDATA
        with two iterations and MPDATA with 44 iterations.
      The colour scale and the wire-frame surface correspond to signal amplitude.
      See section~\ref{sec:example} for discussion.
    }
  \end{figure}
  \includegraphics[height=\textheight]{demo}

>>>>

    \author{~}
%    \author[1]{Sylwester Arabas\footnote{Correspondence to: S. Arabas -- sarabas@igf.fuw.edu.pl -- 0048225546827 -- University of Warsaw, Pasteura 7, 02-093 Warszawa, Poland}}
%    \author[1]{Dorota Jarecka}
%    \author[1]{Anna Jaruga}
%    \author[2]{Maciej Fijałkowski}

%    \affil[1]{Institute of Geophysics, Faculty of Physics, University of Warsaw}
%    \affil[2]{PyPy Team}

@@

  Object oriented programming (OOP) {\em ''has become recognised as the almost unique successful 
    paradigm for creating complex software''} \citep[][Sec.~1.3]{Press_et_al_2007}.
  It is intriguing that, while the quoted statement comes from the very book subtitled 
   {\em The Art of Scientific Computing}, hardly any (if not none) of the currently operational 
    weather and climate prediction systems - flagship examples of complex scientific software - 
    make extensive use of OOP techniques\footnote{
  Fortran has been the language of choice in oceanic \citep{Griffies_et_al_2000}, 
    weather-prediction \citep{Sundberg_2009} and Earth system \citep{Legutke_2012} modelling, 
    and none of its 20\textsuperscript{th}-century editions were object-oriented languages \citep[see e.g.][for discussion]{Norton_et_al_2007}}.

  Application of OOP techniques in development of numerical modelling software may help to:
  \begin{enumerate}[label=(\roman*), leftmargin=*, widest=ii]
    \item{maintain modularity and separation of program logic layers (e.g. separation of
      numerical algorithms, parallelisation mechanisms, data input/output, error handling and
      the description of physical processes); and}
    \item{{\bf shorten and simplify the source code and improve its readability by reproducing within 
      the program logic the mathematical notation used in the literature}.}
  \end{enumerate}
  The first application is attainable, yet arguably cumbersome, with procedural programming.
  The latter, virtually impossible to obtain with procedural programming, is the focus of this paper.
  The importance of reproducing the mathematical notation in the code lays primarily
    in the fact that code readability and brevity 
    %\citep[and not the language choice itself,][]{Seeley_2004}
    significantly contribute to code maintainability \citep{Wilson_et_al_2012}.

@@

  All three programs include an equally structured implementation of the two-dimensional version of the 
    Multidimensional Positive Definite Advective Transport Algorithm \citep[MPDATA;][]{Smolarkiewicz_1984}.
  MPDATA is an example of a numerical procedure used in weather, climate, ocean and solar simulation systems
    \citep[e.g.][respectively]{Ziemianski_et_al_2011,Abiodun_et_al_2011,Ezer_et_al_2002,Charbonneau_and_Smolarkiewicz_2013}.
  The basic MPDATA scheme presented herein is complex enough to contain
    a wide range of mathematical abstractions that can be represented using OOP constructs, 
    yet it is simple enough to allow inclusion of the entire source code within the paper text.
  All relevant MPDATA formul\ae~are given in the text alongside corresponding code fragments
    allowing comparison of the relevant syntax with the mathematical notation.
  These formul\ae~are presented without derivation or detailed discussion
    (see~\citep{Smolarkiewicz_2006} for a recent review of MPDATA-based techniques including an introductory description 
    of the algorithm and an exhaustive list of references).

  The paper is structured as follows.
  In section~\ref{sec:impl} we introduce the ''formula translation'' part of the three implementations
    briefly describing the algorithm itself and
    discussing where and how the OOP techniques were applied in its implementation.
  The remaining part of the implementations -- the solver logic -- is
    presented in appendix~\ref{app:solver}.
  A usage example is given in appendix~\ref{app:usage}
  Section~\ref{sec:perf} covers performance evaluation of the three implementations.
  Section~\ref{sec:tradeoffs} covers discussion of the tradeoffs of the programming language choice.
  Section~\ref{sec:concl} closes the article with a brief summary.

@@

  \section{Implementation of the formul\ae}\label{sec:impl}

@@

  In NumPy, in order to prevent copying of array data during slicing, one needs to operate on the
    so-called array views.
  Array views are obtained when indexing the arrays with objects of the Python's
    built-it \prog{slice} type (or tuples of such objects in case of multi-dimensional arrays).
  Python forbids overloading of operators of built-in types such as \prog{slices}, 
    and does not define addition/subtraction operators for \prog{slice} and \prog{int} pairs.
  Consequently, a custom logic has to be defined not only for fractional indexing,
    but also for shifting the slices by integer intervals ($i\pm1$).
  It is implemented here by declaring a \prog{shift} class with the adequate operator overloads:
  \codepyt{code-pyt-listings.py}{listing03}{listing04}{1}
    and two instances of it to represent unity and half
    in expressions like \prog{i+one}, \prog{i+hlf}, where \prog{i} is an instance of \prog{slice}%
    \footnote{\label{fnt:slice}One could argue that not using an own implementation of a slice-representing class
    in NumPy is a design flaw -- being able to modify behaviour of a hypothetical numpy.slice class 
    through inheritance would allow to implement the same behaviour as obtained in listing P.3 without the need to represent 
    the unity as a separate object}:
  \codepyt{code-pyt-listings.py}{listing04}{listing05}{1}

@@

  Fortran does not feature an analogous mechanism for specifying array ranges in multiple
    dimensions using a single entity.
  As a workaround, in in the Fortran implementation, \prog{pi()} returns a pointer to the array elements specified
    by \prog{i} and \prog{j} interpreted as (i,j) or
    (j,i) depending on the value of the argument \prog{d}.
% TODO: disucss creation of temporary arrays and disabling of vectorisation due to employment of POINTERS
  In addition to \prog{pi()}, a helper \prog{span()} function returning the 
    length of one of the vectors passed as argument is defined:
  \codefor{code-for-listings.f}{listing04}{listing05}{1}
  The \prog{span()} function is used to shorten the declarations of arrays to be returned 
    from functions in the Fortran implementation (see listings F.7 and F.12--F.15).

@@

  The brevity of the code in the above listings as well as its similarity
    to the mathematical notation is the main point of this paper.
  The ''formula translation'' features include:
  \begin{itemize}
    \item{loop-free notation;}
    \item{array-valued functions enabling reuse of sub-expressions;}
    \item{fractional indexing obtained with the help of operator overloading;}
    \item{dimension-independent indexing with the help of \prog{pi} permutation functions.}
  \end{itemize}
  The same features are applied to translation of more complex formul\ae~in the
    following section.

@@

  If the (positive-defined) denominator in equations \ref{eq:A} or \ref{eq:B} 
    equals zero for a given {\em i} and {\em j}, %TODO: >0 in the code
    the corresponding $A_{[i,j]}$ and $B_{[i,j]}$ are set to zero what may be conveniently 
    represented with the \prog{where} construct in all three considered languages:
  \codecpp{code-cpp-listings.hpp}{listing10}{listing11}{1}
  \codepyt{code-pyt-listings.py}{listing09}{listing10}{1}
  \codefor{code-for-listings.f}{listing10}{listing11}{1}
  \codefor{code-for-listings.f}{listing11}{listing12}{1}

@@

  The above listings conclude the formula-translation part of this paper.
  Implementation of a prototype MPDATA solver using the above code
    is presented in appendix~\ref{app:solver}.

@@

  In addition to the standard Python implementation CPython,
    the Python code was tested with PyPy.
  PyPy is an alternative implementation of Python featuring a just-in-time compiler. 
  PyPy includes an experimental partial re-implementation of NumPy that compiles NumPy expressions into native assembler.
  Thanks to employment of lazy evaluation of array expressions (cf. Sect. \ref{sec:array})
    PyPy allows to eliminate the use of temporary matrices for storing intermediate results,
    and to perform multiple operations on the arrays within a single array index traversal
    \footnote{Lazy evaluation available in PyPy 1.9 has been temporarily removed from PyPy during a refactoring of
      the code. It'll be reinstantiated in the codebase as soon as possible, but past PyPy 2.0 release}.
  Consequently, PyPy allows to overcome the same performance-limiting factors as those addressed by Blitz++, although 
    the underlying mechanisms are different.
  In contrast to other solutions for improving performance of NumPy-based codes such as
    Cython\footnote{see \url{http://cython.org}}, 
    numexpr\footnote{see \url{http://code.google.com/p/numexpr/}} or 
    Numba\footnote{see \url{http://numba.pydata.org/}}, 
    PyPy does not require any modifications to the code.
  Thus, PyPy may serve as a drop-in replacement for CPython ready to be used with 
    previously-developed codes.

  The same set of tests was run with all four set-ups.
  Each test set consisted of 16 program runs.
  The test programs are analogous to the example code presented in appendix~\ref{app:usage}.
  The tests were run with different grid sizes ranging from 64$\times$64 to 2048$\times$2048.
  The Gaussian impulse was advected for $nt=2^{24}/(nx\cdot ny)$ timesteps ($2^{24}$ chosen arbitrarily), 
    in order to assure comparable timing accuracy for all grid sizes.
  Three MPDATA iterations were used (i.e. two corrective steps).
  %The initial condition was loaded from a text file, and the final values were compared at the end of the test
  %  with values loaded from another text file assuring the same results were obtained with all four set-ups.
  The tests were run multiple times; program start-up, data loading, and output verification times were
    subtracted from the reported values (see caption of Figure~\ref{fig:cpu-eyrie} for details).

@@

  The CPU time statistics presented in Figures \ref{fig:cpu-eyrie} and~\ref{fig:cpu-skua} reveal
    minor differences between results obtained with the two different processors.
  Presented results lead to the following observations
    (where by referring to language names, only the results obtained with the herein considered
     program codes, and software/hardware configurations are meant):
  \begin{itemize}
    \item{Fortran gives shortest execution times for any domain size;}
    \item{C++ execution times are less than twice those of Fortran for grids larger than 
      256$\times$256;}
    \item{CPython requires from around 4 to almost 10 times more CPU time than Fortran depending on the grid size;}
    \item{PyPy execution times are in most cases closer to C++ than to CPython.}
  \end{itemize}
  The support for OOP features in gfortran, the NumPy support in PyPy, and the relevant optimisation
    mechanisms in GCC are still in active development and hence the performance with some of the set-ups may 
    likely change with newer versions of these packages.

@@

  Three issues specific to Fortran that 
    resulted in employment of a more repetitive or cumbersome syntax than in C++ or Python
    were observed:
  \begin{itemize}
    \item{Fortran lacks support for specifying array ranges in multiple dimensions
      with a single entity (cf. tuples of slices in Python and blitz::RectDomain in C++);
    }
    \item{Fortran does not feature a mechanism allowing to reuse a single piece of code (algorithm)
      with different data types (compare e.g. listings C.15, P.14 and F.17) such as
      templates in C++ and the so-called duck typing in Python;}
    \item{Fortran does not allow function calls to appear on the left hand side
      of assignment (see e.g. how the \prog{ptr} pointers were used as a workaround in the \prog{cyclic\_fill\_halos}
      method in listing F.20);}
    \item{Fortran lacks support for arrays of arrays (cf. Sect. \ref{sec:sequence}).}
  \end{itemize}
  Interestingly, the limitation in extendability via inheritance was found to
    exist partially in NumPy as well (see footnote \ref{fnt:slice}).
  The lack of a counterpart in Fortran to the C++ template mechanism was identified in
    \citep{Cary_et_al_1997}
    as one of the key deficiencies of Fortran when compared with C++ in context 
    of applicability to object-oriented scientific programming.

@@

  Blitz++ ensures temporary-array-free computations by design
    \citep{Veldhuizen_et_al_1997} avoiding unintentional performance loss.
  In contrast, with both Fortran and Python, the memory footprint caused by employment
    of temporary objects in array arithmetics is dependant on compiler choice or
    the level of optimisations.

  \subsection{Developers' community and libraries}%User- and code-base}

@@

  The three languages differ as well with regard to availability of 
    mechanisms for leveraging shared-memory parallelisation (e.g. with multi-core processors).
  GCC supports OpenMP with Fortran and C++.
  The CPython and PyPy implementations of Python do not offer any
    built-in solution for multi-threading.

@@

  The key conclusion is that all considered language/library/compiler
    set-ups offer possibilities for using OOP to compactly 
    represent the mathematical abstractions within the program code. 
  This creates the potential to improve code readability and brevity,
  \begin{itemize}
    \item{contributing to its 
      auditability, indispensable for credible and reproducible research in computational science 
      \citep{Post_et_al_2005, Merali_et_al_2010, Stodden_et_al_2012}; and
    }
    \item{helping to keep the programs maintainable and avoiding accumulation of the code 
      debt\footnote{See \citet{Buschmann_2011} for discussion of technical/code debt.} 
      that besets scientific software in such domains as climate modelling
      \citep{Freeman_et_al_2010}.
    }
  \end{itemize}
  \noindent 
  The performance evaluation revealed that:
    \begin{itemize}
      \item{the Fortran set-up offered shortest execution times,}
      \item{it took the C++ set-up less than twice longer to compute than Fortran,}
      \item{C++ and PyPy set-ups offered significantly smaller memory consumption 
        than Fortran and CPython for larger domains,}
      \item{the PyPy set-up was roughly twice slower than C++ and up to twice faster than CPython.}
    \end{itemize}
  The three equally-structured implementations required ca. 200, 300, and 500 lines of code 
    in Python, C++ and Fortran, respectively.  
  It is the authors' impression that these figures are somehow indicative of the programming 
    effort.

@@

  \section{Prototype solvers}\label{app:solver}

  The following sections provide a complete description of a minimal example of application
    of the formul\ae~''translated'' into C++, Python and Fortran in the main body of the paper.

  \subsection{Halo regions}

@@

  The \prog{advop()} method controls the MPDATA iterations within one timestep.
  The first (step = 0) iteration of MPDATA is an unmodified donor-cell step.
  Subsequent iterations begin with calculation of the antidiffusive Courant fields using 
    formula \ref{eq:antidiff1}.
  In order to calculate values spanning an (i\mhlf~... i+\phlf) range using a formula
    for $C_{[i+\phlf,\ldots]}$ only, the formula is evaluated using extended
    index ranges \prog{im} and \prog{jm}.
  In the second (step=1) iteration the uncorrected Courant field (\prog{C\_unco}) points
    to the original \prog{C} field, and the antidiffusive Courant field is written into
    \prog{C\_corr} which points to \prog{tmp[1]}.
  In the third (step=2) iteration \prog{C\_unco} points to \prog{tmp[1]} while \prog{C\_corr}
    points to \prog{tmp[0]}.
  In subsequent iterations \prog{tmp[0]} and \prog{tmp[1]} are alternately swapped.

@@

  \section{Usage example}\label{app:usage}

@@

  \footnotesize

@@

  \begin{figure}[bh!]
    \caption{\label{fig:mpdata}
      Plot generated by the program given in listing~C.20.
      The top panel shows initial signal shape (at time t=0).
      The subsequent panels show snapshots of the advected field after 75 timesteps
        from three different simulations: donorcell (or 1 MPDATA iteration), MPDATA
        with two iterations and MPDATA with 44 iterations.
      The colour scale and the wire-frame surface correspond to signal amplitude.
      See appendix~\ref{app:usage} for discussion.
    }
  \end{figure}
  \includegraphics[height=\textheight]{demo}
